Hierarchical Region-Network Sparsity for\\ High-Dimensional Inference in Brain Imaging

### 2016-12-06 paper/ipmi2017/paper.tex
finished Bertrands comments

\includegraphics[width=0.45\textwidth]{../figures/level_schematic_crop.pdf}

@@

In sum,
driving sparse model selection by domain knowledge of region-network hierarchies
outcompeted all other frequently used sparse penalization techniques
for high-dimensional data.
%
%
\subsection{Sample complexity of naive versus informed sparse model selection}
% \paragraph{Sample complexity of naive versus informed sparse model selection.}
Subsequently, the sample complexity of
$\ell_1$-penalized and hierarchical-tree-penalized logistic regression
($\alpha = 1, \beta = 1$)
was empirically estimated and quantitatively compared
(Figure \ref{fig_dataratio}).
Region-network priors should constrain model selection towards more
neurobiologically plausible classification estimators.
This should yield better out-of-sample generalization and
support recovery than
neurobiology-naive $\ell_1$-constrained logistic regression
in the data-scarce and data-rich scenarios.
%
The HCP data with examples from 18
psychological tasks were first divided into
90\% of training set (i.e., 7584 neural activity maps) and
10\% of test set (i.e., 842 maps).
Both learning algorithms were fitted based on the
training set at different subsampling fractions:
20\% (1516 neural activity maps),
40\% (3033 maps),
60\% (4550 maps),
80\% (6067 maps), and
100\% (7584 maps).
% results: accuracy
Regarding classification performance on the identical test set,
$\ell_1$-penalized versus hierarchical-tree-penalized logistic regression
achieved
83.6\% versus 88.7\% (20\% of training data),
85.0\% versus 89.2\% (40\%),
86.8\% versus 89.8\% (60\%),
88.9\% versus 90.3\% (80\%),
88.6\% versus 89.7\% (100\%) accuracy.
% results: sparsity
Regarding model sparsity,
the measure $s = \frac{||w||_1}{||w||_F}$ was computed from
the model weights $w$ of both penalized estimators
for each of the 18 classes.
The $\ell_1$-penalized logistic regression
yielded the mean (+/- standard deviation) sparsities
50.0 (+/- 2.6), 45.4 (+/- 2.3), 40.0 (+/- 2.4), 30.9 (+/- 2.0), and 24.0 (+/- 2.3)
after model fitting with 20\% to 100\% training data.
The hierarchical-tree-penalized logistic regression
yield the sparsities
163.2 (+/- 0.7), 160.2 (+/- 1.8), 132.1 (+/- 3.3), 116.2 (+/- 4.4), and 88.4 (+/- 8.4)
after fitting 20\% to 100\% of the training data.
% results: support recovery
Regarding the support recovery, we computed
Pearson correlation $r$ between vectors of
the z-scored model coefficients
the z-scored across-participant average maps for each class.
$\ell_1$-penalized versus hierarchical-tree-penalized logistic regression
achieved a mean correlation $r$ (+/- standard deviation) of
0.10 (+/- 0.03) versus 0.13 (+/- 0.04),
0.11 (+/- 0.03) versus 0.13 (+/- 0.05),
0.13 (+/- 0.04) versus 0.17 (+/- 0.05),
0.16 (+/- 0.04) versus 0.22 (+/- 0.05), and
0.19 (+/- 0.05) versus 0.29 (+/- 0.06) across classes
based on 20\% to 100\% training data.
% results: model variance
Finally, regarding model variance,
we quantified the agreement between
$\ell_1$-penalized versus hierarchical-tree-penalized
model weights after fitting
on 5 different 20\%-subsamples of the training data.
For each classifier,
the absolute model weights were concatenated for all 18 classes,
thresholded at 0.0001 to binarize variable selection,
and mutual information was computed on all pairs
of the 5 trained models.
This agreement metric of model selection across fold pairs
yielded the means 0.001 ($\ell_1$) versus 0.506 (hierarchical tree).
    % \includegraphics[width=0.90\textwidth]{../figures/reg_net_ratio_colin2.pdf}
    \includegraphics[width=0.90\textwidth]{../figures/reg_net_ratio_colin3.png}
    \caption{\textbf{Support recovery as a function of
    region and network emphasis.}
    The relative impact of the region and network priors
    on model selection
    is systematically varied against each other
    (i.e., $\alpha$ and $\beta$ are changed reciprocally).
    Horizontal brain slices are shown with the voxel-wise weights for each class
    from the fitted predictive model.
    The region-network ratio (\textit{columns}) weighted voxel groups
    to priviledge sparse models in function space
    that acknowledge known brain region neighborhoods
    (\textit{left columns}) or
    known brain networks neighborhoods
    (\textit{right columns}).
    Among the 18 classes, the model weights are shown for 3 exemplary
    psychological
    tasks followed by participants lying in a brain imaging scanner
    (\textit{from top to bottom}): tongue movement,
    viewing locations and tools.
    The 18-class out-of-sample accuracy \textit{bottom} and
    the class-wise mean neural activity
    (\textit{rightmost column}, thresholded at the $75^{th}$ percentile)
    are indicated.
    % First,
    Different emphasis on regions versus networks
    in hierarchical structured sparsity can
    yield very similar out-of-sample generalization.
    % Second,
    Favoring region versus network structure during model selection
    recovers complementary, non-identical aspects of
    the neural activity pattern underlying the psychological tasks.
    % Third,
    % equal emphasis on region versus network neighborhoods in the
    % high-dimensional neuroimaging data yields more dispersed,
    % less interpretable model choices.
    \label{fig_regnetratio}
\begin{table*}[]
  \caption{Out-of-sample performance by region-network emphasis}
  \begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
    \textbf{Reg-Net Ratio}          & 100   & 50    & 10   & 5     & 2    & 1     & $\frac{1}{2}$  & $\frac{1}{5}$  & $\frac{1}{10}$ & $\frac{1}{50}$ & $\frac{1}{100}$ \\ \hline
    \textbf{Accuracy {[}\%{]}}             & 89.7  & 89.9  & 90.1 & 90.5  & 88.0 & 89.7  & 87.8           & 88.0           & 87.7           & 88.4           & 88.1  
  \end{tabular}
  \label{table_reg_net_ratio}
\end{table*}

%
Three observations have been made.
First,
in the data-scarce scenario (i.e., 1/5 of available training data),
hierarchical tree sparsity achieved the biggest advantage
in out-of-sample performance by 5.1\% as well as
better support recovery with weight maps already much closer
to the respective class averages
\cite{varoquaux2012small}.
In the case of scarce training data, which is typical for the brain imaging domain,
regularization by region-network priors thus allowed for
more effective extraction of classification-relevant structure
from the neural activity maps.
%
Second,
across training data fractions,
the weight maps from ordinary logistic regression exhibited
higher variance and more zero coefficients
than hierarchical tree logistic regression.
Given the usually high multicollinearity in neuroimaging data,
this observation is likely to reflect instable selection of
representative voxels among class-responsive groups
due to the $\ell_1$-norm penalization.
%
Third,
in the data-rich scenario (i.e., entire training data used for model fitting),
neurobiologically informed logistic regression
profited much more from the increased information quantities than
neurobiologically naive logistic regression.
That is, the region-network priors actually further enhanced the support
recovery in abundant input data.
This was the case although
the maximal classification performance of $\approx$90\% has already
been reached with small training data fractions by the structured estimator.
In contrast, 
the unstructured estimator approached this generalization performance
only with bigger input data quantities.
%
\subsection{Support recovery as a function of region-network emphasis}
% \paragraph{Support recovery as a function of region-network emphasis.}
Finally, the relative importance of the
region and network group penalties within the hierarchical tree prior
was quantified (Figure \ref{fig_regnetratio}).
The group weight $\eta_g$ of region priors was multiplied with a
region-network ratio, while the
group weight $\eta_h$ of network priors was divided by that
region-network ratio. For instance, a region-network ratio of 3
increased the relative importance of known region structure
by multiplying $\beta = \frac{3}{1}$ to
$\eta_g$ of all region group penalties
and multiplying
$\alpha = \frac{1}{3}$ to $\eta_h$ of all network group penalties
(Table \ref{table_reg_net_ratio}).

As the most important observation,
a range between region-dominant and network-dominant structured penalties
yielded quantitatively similar generalization to new data
but qualitatively different decision functions manifested in the weight maps
(Figure \ref{fig_regnetratio}, second and forth column).
Classification models with many zero coefficients but high absolute
coefficients in either region compartments or network compartments
can similarly extrapolate to unseen neural activity maps.
Second,
these achieve classification performance
comparable to equilibrated region-network priors
that set less voxel coefficients to zero and spread the
probability mass across the whole brain
with lower absolute coefficients
(Figure \ref{fig_regnetratio}, third column in the middle).
Third,
overly strong emphasis on either level of the hierarchical prior
provides the neurobiologically informative results with maps
of the most necessary region or network structure for
statistically significant generalization
(Figure \ref{fig_regnetratio}, leftmost and rightmost columns).
%
In sum,
stratifying the hierarchical tree penalty between region and network emphasis
suggests that \textit{class-specific region-network tradeoffs}
enables more performant and more interpretable classification models
for future neuroimaging analyses.

>>>>

\includegraphics[width=0.45\textwidth]{../figures/level_schematic_crop.jpg}

@@

In sum,
driving sparse model selection by domain knowledge of region-network hierarchies
outcompeted all other frequently used sparse penalization techniques
for high-dimensional data.
%
%
\subsection{Sample complexity of naive versus informed sparse model selection}
% \paragraph{Sample complexity of naive versus informed sparse model selection.}
Subsequently, the sample complexity of
$\ell_1$-penalized and hierarchical-tree-penalized logistic regression
($\alpha = 1, \beta = 1$)
was empirically evaluated and quantitatively compared
(Figure \ref{fig_dataratio}).
Region-network priors should constrain model selection towards more
neurobiologically plausible classification estimators.
This should yield better out-of-sample generalization and
support recovery than
neurobiology-naive $\ell_1$-constrained logistic regression
in the data-scarce and data-rich scenarios.
%
The HCP data with examples from 18
psychological tasks were first divided into
90\% of training set (i.e., 7584 neural activity maps) and
10\% of test set (i.e., 842 maps).
Both learning algorithms were fitted based on the
training set at different subsampling fractions:
20\% (1516 neural activity maps),
40\% (3033 maps),
60\% (4550 maps),
80\% (6067 maps), and
100\% (7584 maps).

%
% results: accuracy
Regarding classification performance on the identical test set,
$\ell_1$-penalized versus hierarchical-tree-penalized logistic regression
achieved
83.6\% versus 88.7\% (20\% of training data),
85.0\% versus 89.2\% (40\%),
86.8\% versus 89.8\% (60\%),
88.9\% versus 90.3\% (80\%),
88.6\% versus 89.7\% (100\%) accuracy.
%
% results: sparsity
Regarding model sparsity,
the measure $s = \frac{||w||_1}{||w||_F}$ was computed from
the model weights $w$ of both penalized estimators
for each of the 18 classes.
The $\ell_1$-penalized logistic regression
yielded the mean sparsities
50.0, 45.4, 40.0, 30.9, and 24.0
after model fitting with 20\% to 100\% training data.
The hierarchical-tree-penalized logistic regression
yield the sparsities
163.2, 160.2, 132.1, 116.2, and 88.4
after fitting 20\% to 100\% of the training data.
% results: support recovery
Regarding the support recovery, we computed
Pearson correlation $r$ between vectors of
the z-scored model coefficients
and
the z-scored across-participant average maps for each class.
$\ell_1$-penalized versus hierarchical-tree-penalized logistic regression
achieved a mean correlation $r$ of
0.10 versus 0.13,
0.11 versus 0.13,
0.13 versus 0.17,
0.16 versus 0.22, and
0.19 versus 0.29 across classes
based on 20\% to 100\% training data.
% results: model variance
Finally, regarding model variance,
we quantified the agreement between
$\ell_1$-penalized versus hierarchical-tree-penalized
model weights after fitting
on 5 different 20\%-subsamples of the training data.
For each classifier,
the absolute model weights were concatenated for all 18 classes,
thresholded at 0.0001 to binarize variable selection,
and mutual information was computed on all pairs
of the 5 trained models.
This agreement metric of model selection across fold pairs
yielded the means 0.001 ($\ell_1$) versus 0.506 (hierarchical tree).
\begin{figure*}
  \vspace{-0.6cm}
  \begin{centering}
    % \includegraphics[width=0.90\textwidth]{../figures/reg_net_ratio_colin2.pdf}
    \includegraphics[width=0.90\textwidth]{../figures/reg_net_ratio_colin3.png}
    \vspace{-0.3cm}
    \caption{\textbf{Support recovery as a function of
    region and network emphasis.}
    The relative strength of the region and network priors
    on the regularization
    is systematically varied against each other
    (i.e., $\alpha$ and $\beta$ are changed reciprocally).
    Horizontal brain slices are shown with the voxel-wise weights for each class
    from the fitted predictive model.
    The region-network ratio (\textit{columns}) weighted voxel groups
    to priviledge sparse models in function space
    that acknowledge known brain region neighborhoods
    (\textit{left columns}) or
    known brain networks neighborhoods
    (\textit{right columns}).
    Among the 18 classes, the model weights are shown for 3 exemplary
    psychological
    tasks followed by participants lying in a brain imaging scanner
    (\textit{from top to bottom}): tongue movement,
    viewing locations and tools.
    The 18-class out-of-sample accuracy \textit{bottom} and
    the class-wise mean neural activity
    (\textit{rightmost column}, thresholded at the $75^{th}$ percentile)
    are indicated.
    %
    % First,
    Different emphasis on regions versus networks
    in hierarchical structured sparsity can
    yield very similar out-of-sample generalization.
    %
    % Second,
    Favoring region versus network structure during model selection
    recovers complementary, non-identical aspects of
    the neural activity pattern underlying the psychological tasks.
    %
    % Third,
    % equal emphasis on region versus network neighborhoods in the
    % high-dimensional neuroimaging data yields more dispersed,
    % less interpretable model choices.
    }
    \label{fig_regnetratio}
  \end{centering}
  \vspace{-0.6cm}
\end{figure*}

%
Three observations have been made.
First,
in the data-scarce scenario (i.e., 1/5 of available training data),
hierarchical tree sparsity achieved the biggest advantage
in out-of-sample performance by 5.1\% as well as
better support recovery with weight maps already much closer
to the respective class averages
\cite{varoquaux2012small}.
In the case of scarce training data, which is typical for the brain imaging domain,
regularization by region-network priors thus allowed for
more effective extraction of classification-relevant structure
from the neural activity maps.
\begin{table*}[]
  \vspace{-0.6cm}
  \centering
  \caption{Out-of-sample performance by region-network emphasis}
  \vspace{-0.3cm}
  \begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
    \textbf{Reg-Net Ratio}          & 100   & 50    & 10   & 5     & 2    & 1     & $\frac{1}{2}$  & $\frac{1}{5}$  & $\frac{1}{10}$ & $\frac{1}{50}$ & $\frac{1}{100}$ \\ \hline
    \textbf{Accuracy {[}\%{]}}             & 89.7  & 89.9  & 90.1 & 90.5  & 88.0 & 89.7  & 87.8           & 88.0           & 87.7           & 88.4           & 88.1  
  \end{tabular}
  \vspace{-0.6cm}
  \label{table_reg_net_ratio}
\end{table*}
%
Second,
across training data fractions,
the weight maps from ordinary logistic regression exhibited
higher variance and more zero coefficients
than hierarchical tree logistic regression.
Given the usually high multicollinearity in neuroimaging data,
this observation is likely to reflect instable selection of
representative voxels among class-responsive groups
due to the $\ell_1$-norm penalization.
%
Third,
in the data-rich scenario (i.e., entire training data used for model fitting),
neurobiologically informed logistic regression
profited much more from the increased information quantities than
neurobiologically naive logistic regression.
That is, the region-network priors actually further enhanced the support
recovery in abundant input data.
This was the case although
the maximal classification performance of $\approx$90\% has already
been reached with small training data fractions by the structured estimator.
In contrast, 
the unstructured estimator approached this generalization performance
only with bigger input data quantities.
%
\subsection{Support recovery as a function of region-network emphasis}
% \paragraph{Support recovery as a function of region-network emphasis.}
Finally, the relative importance of the
region and network group penalties within the hierarchical tree prior
was quantified (Figure \ref{fig_regnetratio}).
The group weight $\eta_g$ of region priors was multiplied with a
region-network ratio, while the
group weight $\eta_h$ of network priors was divided by that
region-network ratio. For instance, a region-network ratio of 3
increased the relative importance of known region structure
by multiplying $\beta = \frac{3}{1}$ to
$\eta_g$ of all region group penalties
and multiplying
$\alpha = \frac{1}{3}$ to $\eta_h$ of all network group penalties
(Table \ref{table_reg_net_ratio}).

As the most important observation,
a range between region-dominant and network-dominant structured penalties
yielded quantitatively similar generalization to new data
but qualitatively different decision functions manifested in the weight maps
(Figure \ref{fig_regnetratio}, second and forth column).
Classification models with many zero coefficients but high absolute
coefficients in either region compartments or network compartments
can similarly extrapolate to unseen neural activity maps.
Second,
these achieve classification performance
comparable to equilibrated region-network priors
that set less voxel coefficients to zero and spread the
probability mass across the whole brain
with lower absolute coefficients
(Figure \ref{fig_regnetratio}, third column in the middle).
Third,
overly strong emphasis on either level of the hierarchical prior
provides the neurobiologically informative results with maps
of the most necessary region or network structure for
statistically significant generalization
(Figure \ref{fig_regnetratio}, leftmost and rightmost columns).
%
In sum,
stratifying the hierarchical tree penalty between region and network emphasis
suggests that \textit{class-specific region-network tradeoffs}
enable more performant and more interpretable classification models
for neuroimaging analyses.
